Fluent statistical computing interfaces for biological data analysis

### 2020-10-28 response/response.Rmd

---
title: Thesis Amendments
author: Stuart Lee
output:
  MonashEBSTemplates::memo:
    includes:
      in_header: preamble2.tex
---

@@

I have expanded this section to reflect these comments.

### Chapter 2 comments

Most of the comments pertain to the use and learning of the **plyranges** package. I believe they are best addressed through the **plyranges** documentation and vignette. I have amended the thesis to include the package vignette as an appendix.

### Chapter 3 comments

### Chapter 4 comments

> Much of this chapter depends on understanding what is meant by a coverage trace plot. You do have examples (e.g. Figures 4.4-4.7), but this is at the very end of the chapter. It would be good to provide a motivating figure be added at the start to explain what is meant by “coverage” in this context?
> Pg 63: Introduce what you mean by coverage traces and provide more motiviation why you would want this over just numerical summaries of the data.

> Pg 66: You state “resulting GRanges in memory can be compressed using run-length encoding for any categorical variable”. Could you give a numeric example of the amount of compression that can be obtained with RLEs? Also, maybe add a brief sentence explaining what are RLEs.

> Pg 68: Can you comment on why the GFF files from RefSeq were used?

The model organism is Zebrafish which has limited annotations available compared to human and mouse. I chose the RefSeq GFF build because that is what was recommended to me by our collaborators as the canonical reference.

> Pg 68: How did you handle reads that map over boundaries for introns and exons? or do you only keep the ones that are entirely inside an intron or exon?

Reads that are split are across exon / intron boundaries are counted towards both categories.

> Pg 69: In the facets, you used a short descriptor, but could you write out what is each facet label is in the legend?

I think the facet labels are sufficient in conjunction with the figure captions.

@@

> Pg 69: In the facets, you used a short descriptor, but could you write out what is each facet label is in the legend?

I think the facet labels are sufficient in conjunction with the figure captions.

### Chapter 5 comments

>  One question I had as I was reading throughout this chapter, was what is your recommendation for users who want to calculate distances? can the liminal framework be used to better approximate distances instead of users who perform clustering directly on top of the tSNE or UMAP spaces? While
is not ideal, and it is recommended clustering should performed in the PC space, it happens all the time. Another way of framing this question is can your framework be used to improve distance estimates in the tour with a NLDR method as you note it preserves global structure?

This is a really interesting question. To begin, I have amended the clustering case study to comment on a more realistic approach to perform cluster verification. I like the idea of using the tour to calibrate distances obtained from an NLDR method (or the other way around). As mentioned in the text if there are (well-behaved) clusters in the data, it appears that t-SNE will find them but often messes up their orientation and topology. I think you would get some of the way there using the spatial linking approach I described. Another approach would be to look at distance preservation between successive tour frames. One could then ask, how often are these groups of points close according to the tour? Of interest would also be the manual clustering approaches described in Cook and Swayne, 2007. This is left as future work.

### 2020-10-28 Rmd/04-intron.Rmd
chapter 4 comments

In high-throughput sequencing data sets, coverage is the estimated number of
reads that overlap a single position of the reference genome, and is important for assessing sequencing data quality and used in many different aspects of omics analysis such peak-calling in ChIP-seq or variant calling in DNA-seq [@Sims2014-qk]. Here we emphasise looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data. By faceting these traces over combinations of the experimental design and with their biological context such as gene annotations, we can gain an
insight into biological signal under study. 
Of course, due to the sheer size of most reference genome annotations 
it would take an extremely long time for an analyst to look at every 
single possible region where there is some interesting signal within the
coverage trace. Because visualisation does not scale, we need to search the possible space of coverage traces and provide diagnostics for
identifying traces with interesting biology. We have taken inspiration from
visualisation literature; in particular the idea of scatter plot diagnostics ("scagnostics") for summarising the space of all possible 2D scatter plots
to a small number of descriptors of each scatter plots properties such as 
density or monotonicity [@Friedman2002-yx;@Wilkinson2005-xt].  Similarly,
searching for unusual time series via estimating descriptors such as seasonality or autocorrelation, and visualising those descriptors instead [@Hyndman2015-sc].

@@

The per base coverage score is estimated directly from one or more BAM files
that represent the units within the experimental design,
along with an optional experimental design table that is returns 
a long-form tidy _GRanges_ data structure. The coverage estimation is
computed via the **Rsamtools** package, and
users have the ability to estimate coverage in parallel and drop regions
in the genome where there is no coverage [@rsamtools; @granges]. This
representation is tidy, since each row of the resulting _GRanges_ data
structure corresponds to position(s) within a given sample with a given 
coverage score alongside any variables such as biological group. While the long-form representation repeats the same information
for a sample within the design, the size of the resulting _GRanges_ in
memory can be compressed using run-length encoding for any categorical 
variable. This representation allows us to easily transform the coverage scores
and integrate annotations using the **plyranges** grammar, and visualise traces using **ggplot2** [@Wickham2016-gz].

>>>>

In high-throughput sequencing data sets, coverage is the estimated number of
reads that overlap a single position of the reference genome, and is important for assessing sequencing data quality and used in many different aspects of omics analysis such peak-calling in ChIP-seq or variant calling in DNA-seq [@Sims2014-qk]. Here we emphasise looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data (Figure \@ref(fig:cov-trace). By faceting these traces over combinations of the experimental design and with their biological context such as gene annotations, we can gain an
insight into biological signal under study. 
Of course, due to the sheer size of most reference genome annotations 
it would take an extremely long time for an analyst to look at every 
single possible region where there is some interesting signal within the
coverage trace. Because visualisation does not scale, we need to search the possible space of coverage traces and provide diagnostics for
identifying traces with interesting biology. We have taken inspiration from
visualisation literature; in particular the idea of scatter plot diagnostics ("scagnostics") for summarising the space of all possible 2D scatter plots
to a small number of descriptors of each scatter plots properties such as 
density or monotonicity [@Friedman2002-yx;@Wilkinson2005-xt].  Similarly,
searching for unusual time series via estimating descriptors such as seasonality or autocorrelation, and visualising those descriptors instead [@Hyndman2015-sc].

(ref:example-trace): An illustrative coverage trace plot from Chapter \@ref(ch:plyranges). The coverage _score_ is defined as the number of reads that overlap a single base position within a reference genome. For example, at position 21.807 of chromosome 8 there are approximately 12 reads overlapping it. One could also think of a coverage trace as a histogram with the bin width varies as you move along the genome.

```{r cov-trace, echo = FALSE, fig.cap="(ref:example-trace)}
knitr::include_graphics(here::here("img", "peak-viz-1.png"))
```

@@

```{r rle-example, echo =FALSE}
set.seed(1836)
x <- sample(letters[1:2], 10, replace = TRUE)
x_rle <- S4Vectors::Rle(x)
```
The per base coverage score is estimated directly from one or more BAM files
that represent the units within the experimental design,
along with an optional experimental design table that is returns 
a long-form tidy _GRanges_ data structure. The coverage estimation is
computed via the **Rsamtools** package, and
users have the ability to estimate coverage in parallel and drop regions
in the genome where there is no coverage [@rsamtools; @granges]. This
representation is tidy, since each row of the resulting _GRanges_ data
structure corresponds to position(s) within a given sample with a given 
coverage score alongside any variables such as biological group. While the long-form representation repeats the same information
for a sample within the design, the size of the resulting _GRanges_ in
memory can be compressed using run-length encoding for any categorical 
variable, which is a form of data compression where "runs" of a vector are stored rather than their values. For example, the character vector of letters "a", and "b" is shortened so successive values are stored as a single value with their lengths:

```{r, echo = FALSE}
x
x_rle
```

This representation is memory efficient: at worst it will be the same size as the input, and at best reduce the size by a factor corresponding to the largest run. The allows us to easily transform the coverage scores
and integrate annotations using the **plyranges** grammar, and visualise traces using **ggplot2** [@Wickham2016-gz].
